---
title: "Data model for the schools project"
subtitle: "Planning a star schema"
author: "Clare Gibson"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
# Load knitr package
library(knitr)

# Knitr Options
opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.align = 'center'
)
```

# Introduction
This paper sets out the framework and design principles for building a star schema data model for the schools project.

## Packages
The following packages are required for this analysis.
```{r packages}
library(tidyverse)    # for general wrangling
```


# Assessing the raw data
The raw data for this project is comprised in five dataframes:

* `r_info`
* `r_ks2`
* `r_offers_crit`
* `r_offers_la`
* `r_offers_sch`

## Info
This table has 10 variables and contains characteristics relating to every school in England
```{r head-info}
# Load r_info
load("rda/r_info.rda")

# Inspect the table
r_info %>% 
glimpse()
```

