---
title: "Statistical Exploration of Schools Data"
author: "Clare Gibson"
date: "08/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using primary school performance data to better understand statistical concepts
I intend to use the performance data set that I created in performance.R to revise and better understand the concepts I learned in the edX course [Statistics: Unlocking the World of Data][1] hosted by EdinburghX.

Let's start by loading the libraries we will need.
```{r libraries, results='hide'}
library(tidyverse)
```

Next let's read in the data we need.
```{r read}
load("rda/f_performance.rda")
```

Now we can take a look at the head of the data.
```{r}
head(f_performance)
```
We have the following fields in the data set:

Column Name |Data Type          |Description
------------|-------------------|---------------
lea         |Qualitative Nominal|Local Education Authority (850 = Hampshire, 936 = Surrey)
urn         |Qualitative Nominal|Unique Reference Number of the school
estab       |Qualitative Nominal|Establishment Number of the school
dfe_number  |Qualitative Nominal|DfE Number of the school
laestab     |Qualitative Nominal|LAESTAB Code of the school
schname     |Qualitative Nominal|School name

[1]: <https://learning.edx.org/course/course-v1:EdinburghX+STATSx+1T2021/home> "edX EdinburghX Stats Course"